### 2022-02-03 preprint.tex
Generalizing Scala's Singleton Types

XXXX
In this paper, we report on our attempt at combining an industrial mixed-paradigm language, Scala, with dependent types.
We offer both a formalization of our type system and a discussion of the challenges faced in a practical implementation.
It is our hope that the present paper will serve as a guide for other language implementors interested in pushing the limit of their type systems by adding dependent types.

Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring totality of functions through termination checks.
Instead, our focus is on improving type safety by increasing the expressive power of the type system.

Our system generalizes the traditional notion of singleton type: when the lifted term $t$ contains a non-deterministic choice, the resulting type $\singleton{t}$ denotes the set of values that $t$ could possibly reduce to.
As a result, our type system is capable of type computations by manipulating types which are based on terms, but can nonetheless contain more than a single value.
In combination with subtyping, this allows us to seamlessly integrate with impure, or imprecisely-typed programs.

With these goals in mind, we extended Scala's type system to lift programs and partially evaluate them as part of type-checking.
Users can manipulate dependent types either directly, using new syntactic forms in Scala's type language, or have them inferred automat
ically using an additional keyword.
This effectively allows users to execute programs involving functions and pattern matching to the type level.

We implemented our type system on top of Dotty, the next-generation compiler of Scala.
Working in the context of a production compiler we designed our implementation with backward-compatibility and interoperability in mind
.
Our system is backward-compatible in the sense that all existing programs continue to compile, but also interoperable, in that existing
code can be used from dependently-typed programs and vice versa.
As we will see, subtyping plays a key role in achieving interoperability.

@@

\begin{itemize}
\item
  We begin by motivating why dependent types are desirable and how one might use them to improve type safety (\Cref{sec:motivating-example}). Our example demonstrates how to design a strongly-typed API in a functional style accessible to programmers.
\item
  We describe how we extended Scala with dependent types (\Cref{sec:extending-scala}). This practical introduction would be of interest to any Scala programmer willing to learn how to use our system, as well as programming language designers interested in adding dependent types to their language with subtyping.
\item
  We show a concrete use-case of our system by implementing a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
  Thanks to dependent types, we can statically ensure the type safety of database operations such as join and filter.
  We compare our implementation with an equivalent implicit-based one and show remarkable compilation time savings.
\end{itemize}

Our formalization is out of the scope of this thesis...

@@

We begin by motivating why dependent types are desirable in general purpose programming, and how one might use them to improve type safety.
In our first example, we design an API that keeps track of database tables' schemas in the type.
We demonstrate how dependently-typed list operations can be used to compute schemas resulting from join operations at the type level.
Our second example shows how to build a safer version of the zip operation on lists that only accepts equally-sized arguments.
The examples in this section are written in our dependently-typed extension of Scala described in \Cref{sec:extending-scala}.

@@

\noindent
We can define list concatenation in the usual functional style of Scala\footnote{Our examples use the indentation-based syntax introduced in Scala 3.0.}, that is, using pattern matching and recursion:

@@

\noindent
By annotating a method as |dependent|, the user instructs our system that the result type of |concat| should be as precise as its implementation.
Effectively, this means that the body of |concat| is lifted to the type level, and will be partially evaluated at every call site to compute a precise result type which \emph{depends} on the given inputs.
For recursive |dependent| methods such as |concat|, we infer types that include calls to |concat| itself.
The |<:| annotation lets us provide an upper bound on |concat|'s result type, which will be used while type checking the method's definition.
Finally, by qualifying the definition of |Cons| and |Nil| as |dependent| we also allow their constructors and extractors to be lifted to the type level.
Using these definitions, we can now request the precise type whenever we manipulate lists by annotating the new |val| binding with |dependent|:

@@

Enclosing a pure term in braces (|{|~\ldots{}~|}|) denotes the singleton type of that term.
In the last two lines of this example we are therefore asking our system to prove that |l3| has size 2 and is equivalent to |Cons("A", Cons("B", Nil()))|.

In Scala we often deal with impure or imprecisely-typed code, however.
To integrate with such terms, we provide the |choose[T]| construct.
Operationally, we interpret |choose[T]| as a non-deterministic choice from |T|, which can be modeled faithfully on the type level as an existentially quantified inhabitant of |T| in a singleton type.
Thus, we equate |{ choose[T] }| to |T|, and when typing an impure term such as |Cons(readString(), Nil())| we can assign the precise type |{ Cons(choose[String], Nil()) }|.
Returning to the previous example, this means that even in the presence of impurity, we can perform useful type-level computation and checking:

\begin{lstlisting}
dependent val l2 = Cons(readString(), Nil())
dependent val l3 = l1.concat(l2)
l3: { Cons("A", Cons(choose[String], Nil())) }
\end{lstlisting}

In a style similar to |concat|, we can define remove on |Lst|:

@@

\begin{lstlisting}
dependent val schema1 = Cons("age", Cons("name", Nil()))
dependent val schema2 = Cons("name", Cons("unit", Nil()))
dependent val table1  = Table(schema1, (*\text{\dots}*))
dependent val table2  = Table(schema2, (*\text{\dots}*))
dependent val joined  = table1.join(table2, "name")
joined: { Table(
            Cons("name", Cons("age", Cons("unit", Nil()))),
            choose[DataFrame]) }
\end{lstlisting}

@@

In this section we give an overview of how we extended Scala with dependent types.
This development was an experiment to explore the feasibility of adding dependent types in Scala.
We implemented our prototype as an extension of Dotty, the reference Scala~3 compiler.
Our presentation focuses on several facets of the implementation that are not reflected in our formalism.

On a syntactic level, our Scala extension consists of three additions:

\begin{itemize}
\item
  the singleton types syntax |{ t }|,
\item
  the |dependent| modifier for methods, values and classes,
\item
  the |choose[T]| construct.
\end{itemize}

\noindent
The newly-introduced singleton type syntax enables a subset of Scala expressions to be used in types.
This subset approximately corresponds to the core functional subset of Scala, plus the |choose[T]| construct, as illustrated in \oursystem.
Within this subset, the main differences between our formalism and implementation lie in the handling of pattern matching.

@@

When inferring the result type of a |dependent| method, our system lifts the method's body into a type.
This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |choose[T]|.

@@

\begin{lstlisting}
{ if (personalized) choose[String] else "Joe" }
\end{lstlisting}

@@

In this section, we extend the motivating example presented in \Cref{sec:motivating-example} by building a type-safe interface for Spark datasets.
We use dependent types to implement a simple domain-specific type checker for the SQL-like expressions used in Spark.
We then compare the compilation time of our dependently-typed interface against an equivalent encoding based on implicits.

@@

\begin{lstlisting}
def addRes(a: Any, b: Any) =
  (a, b) match
    case (_: Char, _: Char | Byte | Int  | Long | Double) => choose[Double]
    case (_: Byte, _: Byte | Int  | Long | Double)        => b
    case (_: Int,  _: Int  | Long | Double)               => b
    case (_: Long, _: Long | Double)                      => b
    case (_: Double, _: Double)                           => choose[Double]
    case (_: Byte | Int | Long | Double, _)               => addRes(b, a)
    case _ => throw new Error("incompatible types in addition")
type AddRes[A, B] = { addRes(choose[A], choose[B]) }
\end{lstlisting}

@@

\begin{lstlisting}
val pred =    table.col("a")    +    table.col("b")    ===   table.col("c")
// Infers: { Column[Int]("a") }   { Column[Long]("b") }   { Column[Long]("c") }
\end{lstlisting}
% val pred = (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("a") \phantom{.} }*) + (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("b") \phantom{.}\phantom{.} }*) === (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("c") \phantom{.}\phantom{.} }*)
% // Infers: (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Int]("a") \}}}*)   (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("b") \}}}*)     (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("c") \}}}*)

@@

In the context of Haskell, Eisenberg's work on Dependent Haskell~\citep{eisenberg2016dependent} is closest to ours, in that it adds first-class support for dependent types to an established language, in a backwards-compatible way.
Dependent Haskell supports general recursion without termination checks, which makes it less suitable for theorem proving.
While we share similar goals, our work is differentiated by the contrasting paradigms of Scala and Haskell.
Like many object-oriented languages, Scala is primarily built around subtyping and does not restrict the use of side effects.
Furthermore, Eisenberg's system provides control over the relevance of values and type parameters.
In contrast, our system does not support any erasure annotations and simply follows Scala's canonical erasure strategy: types are systematically erased to JVM types, and terms are left untouched.
Weirich established a fully mechanized type safety proof for the core of Dependent Haskell using the Coq proof assistant~\citep{weirich2017a}.

>>>>

In this paper, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form a dependently-typed programming.
Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring totality of functions through termination checks.
Instead, our focus is on improving type safety by increasing the expressive power of the type system.

TODO: reformulate
Our system generalizes the traditional notion of singleton type: when the lifted term $t$ contains a non-deterministic choice, the resulting type $\singleton{t}$ denotes the set of values that $t$ could possibly reduce to.
As a result, our type system is capable of type computations by manipulating types which are based on terms, but can nonetheless contain more than a single value.
In combination with subtyping, this allows us to seamlessly integrate with impure, or imprecisely-typed programs.

With these goals in mind, we extended Scala's type system to lift programs and partially evaluate them as part of type-checking.
Users can manipulate dependent types either directly, using new syntactic forms in Scala's type language, or have them inferred automatically using an new language keyword.
This effectively allows users to execute programs involving functions and pattern matching to the type level.

@@

\begin{itemize}
\item
  We begin by motivating why type-level programming in desirable and how one might use our Scala extension to improve type safety (\Cref{sec:motivating-example}). Our example demonstrates how to design a strongly-typed API in a functional style accessible to programmers.
\item
  TODO
  We prototyped our type system on top of Dotty, the reference Scala~3 compiler.
  We describe how we extended Scala with dependent types.
  This development was an experiment to explore the feasibility of adding dependent types in Scala.
  We implemented our prototype as an extension of Dotty, the reference Scala~3 compiler.
  Our presentation focuses on several facets of the implementation that are not reflected in our formalism.

  We describe how we extended Scala with dependent types (\Cref{sec:extending-scala}). This practical introduction would be of interest to any Scala programmer willing to learn how to use our system, as well as programming language designers interested in adding dependent types to their language with subtyping.
\item
  We show a concrete use-case of our system by implementing a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
  Thanks to our generalized singleton types, we can statically ensure the type safety of database operations such as join and filter.
  TODO: keep this?
  We compare our implementation with an equivalent implicit-based one and show remarkable compilation time savings.
\end{itemize}

TODO: pretend that formalization doesn't exist?
The said report covers more material than what is included in this chapter, in particular, our formalization and the associated metatheory are out of the scope of this dissertation \citep[Section 3 and 4]{schmid2020coming}.

@@

We begin by motivating why type-level programming is desirable in general purpose programming.
In our first example, we design an API that keeps track of database tables' schemas in the type, and uses that information to improve type safety.
The examples in this section are written in our Scala extension described in \Cref{sec:extending-scala}.

@@

\noindent
We can define list concatenation in the usual functional style of Scala, that is, using pattern matching and recursion:

@@

\noindent
By annotating a method as |dependent|, the user instructs our system that the result type of |concat| should be as precise as its implementation.
Effectively, this means that the body of |concat| is lifted to the type level in a singleton type, and will be partially evaluated at every call site to compute a precise result type which \emph{depends} on the given inputs.
For recursive |dependent| methods such as |concat|, we infer types that include calls to |concat| itself.
The |<:| annotation lets us provide an upper bound on |concat|'s result type, which will be used while type checking the method's definition.
Finally, by qualifying the definition of |Cons| and |Nil| as |dependent| we also allow their constructors and extractors to be lifted to the type level.
Using these definitions, we can now request the precise type whenever we manipulate lists by annotating |val| bindings as |dependent|:

@@

Enclosing a pure term in braces (|{|~\ldots{}~|}|) denotes the singleton type of that term.
In the last two lines of this example, we are therefore asking our system to prove that |l3| has size 2 and is equivalent to |Cons("A", Cons("B", Nil()))|.
Similarly, we can define remove on |Lst|:

@@

\begin{lstlisting}
dependent val schema1 = Cons("age", Cons("name", Nil()))
dependent val schema2 = Cons("name", Cons("unit", Nil()))
dependent val table1  = Table(schema1, (*\text{\dots}*))
dependent val table2  = Table(schema2, (*\text{\dots}*))
dependent val joined  = table1.join(table2, "name")
joined: { Table(
            Cons("name", Cons("age", Cons("unit", Nil()))),
            _: DataFrame) }
\end{lstlisting}

@@

In this section we give an overview of how we extended Scala with a generalized notion of singleton types, and how we manipulate these types during type-checking.
Our presentation is divided into three parts.
First, we introduce new types that reflect a subset of Scala's term language at the type level.
We then introduce the |dependent| qualifier, which influences type inference to assign generalized singleton types to the annotated definition.
Finally, we give an informal description of our type evaluation algorithm.

This development was an experiment to explore the feasibility of adding a lightweight form of dependent types to Scala.
We implemented our prototype as an extension of Dotty, the reference Scala~3 compiler.

\subsection{Reflecting terms in types}

The fundamental difference between our and Scala's existing type system is that we take Scala's notion of singleton types, that is, precise types for variable bindings and literals, and extend them to cover Scala's core functional expressions.
Concretely we add new forms of types for the following constructs:

\begin{itemize}
  \item Variable bindings and member selections
  \item Primitive literals
  \item Method calls
  \item If-then-else expressions
  \item Pattern matching expressions
  \item Constructor calls
\end{itemize}

Such new types can be expressed using the |{| $e$ |}| type syntax, where $e$ is an expression in the core functional subset of Scala listed above.
For instance, suppose $\text{foo}$ stands for the expression

\begin{lstlisting}
if (x % 2 == 0) "even" else "odd"
\end{lstlisting}

\noindent then $\text{foo}$ can be typed as |{| $\text{foo}$ |}|. More precisely, |{| $\text{foo}$ |}| is the singleton type of $\text{foo}$ and corresponds to the unique value of $\text{foo}$ in a given context.

The first two constructs in this list have antecedents in Scala. Types for bindings are available since the early days of Scala with the |x.type| syntax \citep[Section 3.2.1]{scalaSpec}, which is equivalent to |{ x }| in our system.
Types for primitive literals have recently been added to the language: literals for booleans, strings and the various numeric types are made available in the type language \citep{sip23}.

In addition to base types and singleton types for every pure term, our system also supports types that lie in-between. We add a new form, |_: T|, which supplements expressions in types of form |{| $e$ |}| by allowing the user to intersperse terms and base types.

For instance, consider the following list and its singleton type: |Cons(Nil, Nil): { Cons(Nil, Nil) }|.
In our system, this term may also be typed less precisely, such as:

\begin{itemize}
  \item |Cons(Nil, Nil): { Cons(_: Any, Nil   ) }|
  \item |Cons(Nil, Nil): { Cons(Nil   , _: Lst) }|
  \item |Cons(Nil, Nil): { Cons(_: Any, _: Lst) }|
\end{itemize}

all of which are subsumed by |Cons(Nil, Nil): Lst|.

@@

When inferring the result type of a |dependent| method, our system lifts the method's body into a generalized singleton type.
This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|.

@@

\begin{lstlisting}
{ if (personalized) (_: String) else "Joe" }
\end{lstlisting}

@@

In this section, we extend the motivating example presented in \Cref{sec:motivating-example} by building a type-safe interface for Spark datasets.
We use our Scala extension to implement a simple domain-specific type checker for the SQL-like expressions used in Spark.

@@

\begin{lstlisting}
def addRes(a: Any, b: Any) =
  (a, b) match
    case (_: Char, _: Char | Byte | Int  | Long | Double) => _: Double
    case (_: Byte, _: Byte | Int  | Long | Double)        => b
    case (_: Int,  _: Int  | Long | Double)               => b
    case (_: Long, _: Long | Double)                      => b
    case (_: Double, _: Double)                           => _: Double
    case (_: Byte | Int | Long | Double, _)               => addRes(b, a)
    case _ => throw new Error("incompatible types in addition")
type AddRes[A, B] = { addRes(_: A, _: B) }
\end{lstlisting}

@@

\begin{lstlisting}
val pred =    table.col("a")    +    table.col("b")    ===   table.col("c")
// Infers: { Column[Int]("a") }   { Column[Long]("b") }   { Column[Long]("c") }
\end{lstlisting}

@@

In the context of Haskell, Eisenberg's work on Dependent Haskell~\citep{eisenberg2016dependent} is closest to ours, in that it adds first-class support for dependent types to an established language, in a backwards-compatible way.
Dependent Haskell supports general recursion without termination checks, which makes it less suitable for theorem proving.
While we share similar goals, our work is differentiated by the contrasting paradigms of Scala and Haskell.
Like many object-oriented languages, Scala is primarily built around subtyping and does not restrict the use of side effects.
Furthermore, Eisenberg's system provides control over the relevance of values and type parameters.
In contrast, our system does not support any erasure annotations and simply follows Scala's canonical erasure strategy: types are systematically erased to JVM types, and terms are left untouched.
\citeauthor{weirich2017a} established a fully mechanized type safety proof for the core of Dependent Haskell~\citep{weirich2017a}.
